MetaTracts - A Method for Robust Extraction and Visualization of Carbon Fiber Bundles in Fiber Reinforced Composites

### 2014-12-11 chapters/MetaTracts.tex
Major changes fits in 8 pages

\section{Introduction and Motivation}
\label{sec:intro}
Modern industry is increasingly demanding function orientation, integration, and efficiency of novel materials and components. The material of choice for a growing number of applications is carbon fiber reinforced polymer (CFRP), which allows an integration of these continuously rising demands and increasingly replaces conventional materials such as aluminum or steel. 
%CFRPs have been successfully introduced in aeronautic and automotive %applications within recent years, replacing parts made of conventional %materials such as aluminum or steel. But the application areas of CFRPs %are not limited on these two industries. For a wide range of different %areas ranging from marine and civil engineering, wind energy, to sporting %goods manufacturing, CFRP is highly interesting. 
%The main reason therefore is found in CFRP materials having desirable %characteristics such as high specific stiffness, high specific strength %and high corrosion resistance. 
CFRP materials have desirable characteristics such as high specific stiffness, high specific strength and high corrosion resistance. Moreover, CFRP materials show these characteristics at considerably lower weight. At the same time, highly complex and integrated components, which were previously impossible to manufacture may be produced from CFRPs. Primary structures and highly loaded components in aeronautics are one example.
Typically, carbon fiber reinforced polymer components and more specifically CFRP laminates with endless carbon fibers as addressed in this work consist of two main components:
a matrix, which acts as a bonding component, and the reinforcements, which allow for achieving the desired characteristics. Various production processes are used to manufacture CFRP laminates. Most of these processes start with the reinforcement component, weaving individual carbon fiber bundles (yarn) into sheets of carbon fiber cloth in a predefined pattern. These sheets of woven carbon fiber cloth are also referred to as fabric and form the geometrical structure of the final CFRP materials. Depending on the requirements of the final component, fabrics may be stacked in multiple layers in similar or different orientation. Both alignment of fabrics and the weaving pattern of the individual carbon fiber bundles strongly influence the properties of the CFRP laminate. Resins are then integrated in the material system to fill the gaps in the fabric forming the matrix component. The main function of the matrix is to act as a bonding between the individual carbon fiber bundles. After curing, the production process of the CFRP laminate is finished.

The increasing share of CFRPs as well as the complexity of both material system and final component has generated a strong demand towards non-destructive testing (NDT) techniques for quality control~\cite{Red2012}. Ultrasonic testing (UT) is the most commonly used method for this purpose. While UT provides a quick and cost-efficient overview of the material, it also lacks resolution and may generate arbitrary results, e.g., due to the geometry of the component. Industrial 3D X-ray computed tomography (XCT, also referred to as 3DXCT or cone beam XCT) is increasingly applied for non-destructive testing of fiber reinforced polymers~\cite{Kastner2012}. In contrast to UT, XCT generates a highly detailed 3D volumetric representation of the scanned specimen. In cone beam XCT geometry the specimen is placed on a rotary table between X-ray source and detector. The X-rays passing through the specimen get attenuated by the materials present. By transferring the X-rays in a scintillator layer into visible light, the detector records the corresponding 2D attenuation image (penetration image). The specimen is rotated recording at each angular step a 2D attenuation image to generate the full series of attenuation images of a $360^\circ$ rotation required for a complete reconstruction of the data volume ~\cite{heinzl-2008-thesis}. 
While cone beam XCT can reach voxel sizes below 500 $nm$ generating high resolution volume data for comprehensive and detailed analyses, unfortunately there is still a trade-off between viewport and image resolution. The magnification reached within an XCT scan is determined by the used distances between source and specimen as well as source and detector. The magnification therefore directly influences both resolution and viewport: while higher resolutions decrease the viewport but show more details, lower resolutions allow for larger viewports and thus larger portions of the specimen.

In this work, we focus on datasets with larger viewports but lower resolutions where the individual carbon fibers (filaments) are indiscernible or barely visible. Our domain experts are mainly interested in visualizing the geometric structures in the weaving pattern of fiber bundles in endless carbon fiber reinforced composites instead of high resolution studies of individual fibers.
%of short fiber reinforced composites as presented by Weissenbock2014 et al.~\cite{Weissenbock2014}.
Figure~\ref{fig:data-char} depicts our targeted dataset type. It shows the recurring fiber bundle pattern in the final CFRP laminate, the \textit{unit cell}.
Our work is motivated from the recent progress in two interrelated fields: Firstly, CFRP components have gained wide application in industry because of its superior material and physical properties in comparison to conventional materials~\cite{Karpat2012}. Secondly, recent developments of industrial 3D X-ray computed tomography (XCT) with regard to larger detectors, larger field of views, and better resolutions opened XCT for this new application area of NDT for fiber reinforced components~\cite{Schilling2005}. 
%
% Fiber bundles determine to a large extent the properties in terms of specific stiffness, specific strength, or durability of the final component. However, when scanning reinforced polymers using modern XCT devices, there is still a trade-off between view port and image resolution. Due to the intrinsic concept of cone beam and fan beam XCT setups, the magnification reached within a scan is determined by the used distances between source and specimen as well as source and detector. Furthermore the magnification directly influences both resolution and viewport: Higher resolutions decrease the viewport or how much of the structure can be imaged while lower resolutions allow for larger view ports.
%\begin{figure}
%\begin{center}
%\includegraphics[width=0.3\textwidth]{imagesMetaTracts/crop-13-total}
%\caption{Volume rendering of a part of a dataset. Fig~\ref{fig:slice_data:a} and ~\ref{fig:slice_data:b} show image slices along the blue and red plane respectively.  }\label{fig:orig_data}
%\end{center}
%\end{figure}
%!!!!! TODO REFINE DTI PARAGRAPH !!!!!
%Diffusion Tensor Imaging tractography clustering and visualization.
%The topic of DTI visualization has seen an extensive amount of investigation by researchers.
%Advantage of two phase clustering 
%What is the  motivation for fiber-lets
%Seed point selection 
%distance between tracks 
%how well do we do in noise , may be data set 1
%
%!!!!! I have replaced this with the following para !!!!!
%In this work, we present a framework which interprets and advances these techniques in order to extract and visualize fiber bundles in industrial 3D X-Ray computed tomography data of woven fiber reinforced composites. The main goal of this work is to expand the state of the art in nondestructive testing of composite structures towards lower resolutions and thus larger areas of woven fabrics.
%
%While the importance of fiber bundles in determining the final properties of the components have  become evident,  the tools for visualization of the same have not developed at the same pace.
%
While fiber bundles are now understood as highly important in determining component properties, the tools for visualizing the internal structure have not developed at the same pace.
To the best of our knowledge, there is no current work that can resolve simple queries such as:
\begin{itemize}[noitemsep]
\item{ How to extract and visualize the geometric structure of a particular fiber bundle?}
\item{ How to visualize the interaction between a particular pair of fiber bundles (weaving/braiding) or a unit cell?}
\item{ Which fiber bundles show a particular orientation? }
\item{ Which fiber bundles are of the same type of yarn? (i.e. which bundles show similar sizes or diameters, which is the largest or smallest fiber bundle?)}
\end{itemize}

We demarcate the above queries into two parts: \textit{geometric structure} and \textit{spatial context}. Geometric structure refers to the shape, size, and orientation of a fiber bundle or a group of them. Spatial context refers to how two or more bundles interact with each other. It provides answers to questions such as: Are these bundles in contact at a particular position in the data? What is the relative orientations of the contacting bundles? 
Providing answers to the above queries from the volume renderings of XCT datasets or from visual inspection of particular 2D slices is non-trivial, even for experts. We present and evaluate a method which uses visualization techniques to gain insight into our data.
We interpret and advance techniques from diffusion tensor imaging to extract and visualize geometric structures from XCT data of the woven carbon fiber reinforced composites. The main goal of this work is to expand the state of the art in non-destructive testing through visualization of composite structures in \textit{complete unit cells} of woven fabrics.
\begin{figure}[tb]
	\centering
	%\includegraphics[width=0.45\textwidth]{imagesMT2014/algo.png}
	%\includegraphics[width=0.5\textwidth,  trim = 0mm 300mm 0mm 0mm, clip,]{imagesMT2014/image_algo}
	\includegraphics[width=\linewidth]{images_pvis/workflow.pdf}
	\caption{Flow chart of the MetaTracts approach for fiber bundle extraction}
	\label{fig:flowchart}
\end{figure}
Figure~\ref{fig:flowchart} shows an over view of our approach. Starting from XCT data, we first generate MetaTracts (see Sec.~\ref{sec:approach}). Then we cluster the MetaTracts to generate and visualize fiber bundles (see Sec.~\ref{subsec:fiber-bundles}). We discuss the visualizations we created for our domain experts in Sec.~\ref{sec:vis}. Sec.~\ref{sec:results} describes our experimental results, Sec.~\ref{sec:param_choices} presents our parameter choices and  Sec.~\ref{sec:user_eval} finally summarizes the user evaluation of our method.

@@

Diffusion Magnetic Resonance Imagining (dMRI, also referred to as Diffusion Tensor Imaging (DTI)) has gained popularity in medical diagnosis within recent years. Its main clinical application is found in the study and treatment of neurological disorders. DTI may reveal abnormalities in white matter fiber structure and is used for visualizing the organization of fibers in the human brain and brain connectivity.
A variety of algorithms have been proposed for generating fiber-tract trajectories. In general, these reconstructions of fiber trajectories are clustered in bundles which are expected to be related anatomically or spatially. 
We broadly divide the related work on DTI into two parts of immediate relevance to our proposed solution: fiber tracking and fiber clustering. We discuss methods for analyzing the second order local structure and review the current state of the art in the visual analysis of fiber reinforced composites.

\subsection {Fiber Tracking in Diffusion Tensor Imaging}
\label{subsec:fiberEx} 
A basic assumption in DTI analysis is that the principal eigenvector of the diffusion tensor is parallel to the underlying dominant fiber direction in each image voxel~\cite{Basser2002, Basser2000, Mori1999, Mori2002}. Continuous tracts are created by propagating virtual particles along fiber directions until they reach some termination criterion. 
This is usually done by solving a Runge-Kutta integration (typically second or fourth order). The principal diffusion direction at each discrete location is interpolated to form a continuous velocity field. 
Because decisions are made locally, these methods perform poorly in noisy regions and often generate small fibers. Basser et al.~\cite{Basser2002,Basser2000} proposed that white matter tracts could be represented as 3D curves in space. They showed that numerical methods could be used to follow fibers and fiber bundles and to generate tracts in human brain data. 
Mori et al.~\cite{Mori1999,Mori2002} divided reconstruction techniques into line-propagation or energy minimization techniques. In line propagation approaches, trajectories are computed based on local neighborhoods and in energy minimization approaches the most  favorable trajectory connecting two given endpoints is selected.

\subsection {Fiber Clustering in Diffusion Tensor Imaging}
\label {subsec:fiberClus}
In DTI, a similarity measure based on factors such as proximity between fibers is used to group fiber tracts into bundles. Extensive research has been done on automatic DTI fiber clustering methods \cite{Brun2004,Brun2003,Corouge2004,westinMEDIA02,Zhang2008}. 
Clustering requires choosing a suitable proximity measure and a method for grouping ``close" fibers.
Pairwise proximity measures include endpoint distances~\cite{Brun2003} and mean of the closest distances between points on two fibers~\cite{Corouge2004}. Zhang et al.~\cite{Zhang2008} introduced a thresholded version of the of mean of closest distances, so that fibers which are close for certain distance and then diverge, are clustered separately. Brun et al.~\cite{Brun2004} use normalized cuts along with a pairwise distances measure computed using a 9D fiber shape descriptors. The choice of the clustering algorithm can be broadly divided into those approaches using hierarchical clustering~\cite{Moberts2005, Zhang2008} and those using spectral clustering~\cite{Brun2004,jonasson2005, ODonnell2007}.
Brun et al.~\cite{Brun2003} described how a spectral non-linear dimensionality reduction technique, Laplacian eigenmaps (Belkin and Niyogi~\cite{Belkin01}), can be applied to the problem of organizing fiber tracts data. The key notion of the Laplacian eigenmaps algorithm is to represent the underlying data as a graph. Each node represents a data point and the edges connect neighboring data points. An eigenvalue problem is solved to represent the data in a lower dimensional space while preserving the local graph structure. In the case of fiber bundles, the individual points are fiber tracts. In the ideal case fiber tracts belonging to the same bundle must remain ``close" to each other in the lower dimensional space. Westin et al.~\cite{westinMEDIA02} also use spectral clustering on a Hausdorff distance measure defined as the maximum of point wise minimum distances between two fibers. Jonasson et al.~\cite{jonasson2005} run k-means clustering on the eigenvectors of the affinity matrix defined as the co-occurrence of fibers in the data.
%The clustering methods have also been evaluated to be generally accurate in capturing fiber bundles as seen in Moberts et al.~\cite{Moberts2005} and Corouge et al.~\cite{Corouge2004}. 
The agglomerative hierarchical clustering method~\cite{DudaHartStork01} has gained popularity for proximity based fiber segregation (Zhang et al.~\cite{Zhang2008}, Corouge et al.~\cite{Corouge2004}). 
These approaches build on the assumption that proximity measures that compare DTI fiber trajectories can also represent anatomical relationships. An agglomerative hierarchical clustering method starts with each data point/fiber in an individual cluster. At each stage of the algorithm the two most similar clusters are merged based on some criterion. The two basic cluster similarity measures are single-link and complete-link. With the single link measure the distance between the clusters is the distance between the closest pair of items. 
Moberts et al.~\cite{Moberts2005} implemented several distance measures in their evaluation of fiber clustering and concluded that clustering methods are generally accurate in capturing fiber bundles. 
%Mean of closest distances performs better than closest point distance computations or Hausdorff distance measures and hierarchical clustering using single-link measure gave the best results. 
There are a number of difficulties in hierarchical clustering. First, computing all pairs' distances for tracts to generate the distance matrix is time consuming~\cite{Garyfallidis2012}. Second, a ``correct'' distance measure to compare tracts must be chosen. Third, hierarchical clustering is best suited for similar length fibers.
Spectral methods are also hindered by long matrix computations.

\subsection{Second order local structure}
Unlike DTI, we do not have diffusion tensor data. Instead, we have a scalar volume with tubular structures embedded in them. Analyzing curvilinear structures in volumetric images has been utilized for a variety of purposes including center line extraction~\cite{Bouix2005} and vascular image enhancement as proposed by Frangi et al.~\cite{Frangi1998} and Sato et al.~\cite{Sato1997}. Frangi et al.~\cite{Frangi1998} introduced a method based on studying the eigenvalues of the Hessian matrix specifically for the purposes of developing vessel enhancement filters.

\subsection {Visual Analysis and Modeling of Fiber Reinforced Composites}
The approaches presented in visualization and analysis of composites mainly focus on individual objects such as fibers or pores and deals with fiber extraction from high resolution data where the individual fibers are clearly discernible. Fritz et al.~\cite{Fritz2009} proposed workflows for NDT practitioners to explore and quantify steel fibers in reinforced sprayed concrete. This approach allows analyzing fiber orientations based on direction transfer functions. Salaberger et al.~\cite{Salaberger2011} introduced a pipeline to extract and characterize individual fibers of fiber reinforced composites. They encode the extracted fibers as color-coded line segments in 3D and visually identifying fibers with similar orientations. Recently, Weissenbock et al.~\cite{Weissenbock2014} introduced a system for interactive exploration and analysis of fibers in fiber reinforced polymers. 
Lomov et al.~\cite{Lomov2010, Lomov2001} and Straumit et al.~\cite{Straumit2014}, discussed problems and current available solutions in geometric modeling of three dimensional composites. Modeling of the composites, first starts with establishing the topology of the structure, which translates to answering if a particular bundle is in contact with another at a particular position. The second step builds the geometry of the model, answering questions relating to placement of bundles in space their orientations and dimensions. 
%Tools based on image analysis, fiber tracking and clustering into bundles in DTI and their subsequent visualization through stream tubes and coloring schemes have received extensive research. Recently, similar techniques have been applied to fiber tracing in composite materials. 
To the best of our knowledge there is no approach focusing on direct extraction of fiber bundle structures from low resolution XCT of CFRP laminates.

@@

\section {Extracting MetaTracts }
\label{sec:approach}
Extracting the MetaTracts consists of two  major steps. First, we extract a local orientation vector at each grid location (Sec.~\ref{subsec:rh}). Second, we compute coarse poly-cylinders called MetaTracts which traverse the data (Sec.~\ref{subsec:mtprop}).

@@

The input to the first stage of our pipeline is the original scalar dataset in a uniform lattice grid in $\mathbb{R}^3$. As output, each grid location gets associated with an unit vector representing the local orientation and a real value [0,1] representing a measure of the reliability of the local orientation. We approximate the local orientation by eigenvalue analysis of the Hessian matrix computed at each voxel.
The Hessian matrix represents the local curvature. The eigen decomposition of the Hessian matrix gives the eigenvectors which describe the local second-order structure of the image.
The eigenvector corresponding to the smallest eigenvalue (also referred as the principal direction) gives the direction along which the curvature is smallest. This direction coincides with the direction of the tubular structure.

Frangi et al.~\cite{Frangi1998} introduced a process that searches for geometric structures which are tubular. They define a measure based on two geometric ratios of the second order ellipsoid given by the local Hessian matrix to measure the \textit{vesselness} criterion. In order to determine \textit{reliable Hessians} for our approach, we make use of the vesselness criterion as follows: Let $\lambda_{K}$ be the eigenvalue with the $K^{th}$ smallest magnitude. Here, $|{\lambda}_{1}| \leq| {\lambda}_{2}|\leq| {\lambda}_{3}| $ are the eigenvalues of the Hessian matrix. Specifically, a pixel belonging to a vessel region will have small $\lambda_{1}$ ($|\lambda_{1}|\approx 0$) and $\lambda_{2}$, $\lambda_{3}$ of large magnitude and of equal sign($|\lambda_{1}| \ll |\lambda_{2}|$ and $|\lambda_{2}|\approx |\lambda_{3}|$). The sign indicates if the vessel is bright in a dark background or dark in a bright background. In our case the individual fibers are bright ($\lambda_2,\lambda_3 < 0$). The following measures are defined in~\cite{Frangi1998}.

>>>>

\section{Introduction and Motivation}
\label{sec:intro}
Modern industry is increasingly demanding function orientation, integration, and efficiency of novel materials and components. The material of choice for a growing number of applications are carbon fiber reinforced polymers (CFRP), which allow an integration of these continuously rising demands. 
%CFRPs have been successfully introduced in aeronautic and automotive %applications within recent years, replacing parts made of conventional %materials such as aluminum or steel. But the application areas of CFRPs %are not limited on these two industries. For a wide range of different %areas ranging from marine and civil engineering, wind energy, to sporting %goods manufacturing, CFRP is highly interesting. 
%The main reason therefore is found in CFRP materials having desirable %characteristics such as high specific stiffness, high specific strength %and high corrosion resistance. 
CFRP materials are increasingly replacing conventional materials such as aluminum or steel, as they have desirable characteristics such as high specific stiffness, high specific strength and high corrosion resistance. 
Moreover, CFRP materials show these characteristics at considerably lower weight. At the same time, highly complex and integrated components, which were previously impossible to manufacture may be produced from CFRPs. Primary structures and highly loaded components in aeronautics are one example.

Typically, carbon fiber reinforced polymer components and more specifically CFRP laminates with endless carbon fibers as addressed in this work, consist of two main components:
a matrix, which acts as a bonding component, and the reinforcements, which allow for achieving the desired characteristics. Various production processes are used to manufacture CFRP laminates. Most of these processes start with the reinforcement component by weaving individual carbon fiber bundles (yarn) into sheets of carbon fiber cloth in a predefined pattern. These sheets of woven carbon fiber cloth are also referred to as fabric and form the geometrical structure of the final CFRP materials. Depending on the requirements of the final component as well as on the manufacturing process, fabrics may be stacked in multiple layers of the cloth in similar or different orientation. Both alignment of fabrics and the weaving pattern of the individual carbon fiber bundles strongly influence the properties of the CFRP laminate. Resins are integrated in the material system to fill the gaps in the fabric, forming the matrix component. The main function of the matrix is to act as a bonding between the individual carbon fiber bundles. After curing, the production process of the CFRP laminate is finished.

The increasing share of CFRPs as well as the complexity of material system and final components has generated a strong demand towards non-destructive testing (NDT) techniques for quality control~\cite{Red2012}. Ultrasonic testing (UT) is the most commonly used method for this purpose. While UT provides a quick and cost-efficient overview of the material, it also shows low resolution and may generate arbitrary results, e.g. due to the geometry of the final component. Industrial 3D X-ray computed tomography (XCT, also referred to as 3DXCT or cone beam XCT) is increasingly applied for non-destructive testing of fiber reinforced polymers~\cite{Kastner2012}. In contrast to UT, XCT generates a highly detailed 3D volumetric representation of the scanned specimen. In cone beam XCT geometry the specimen is placed on a rotary table between X-ray source and detector. The X-rays passing through the specimen get attenuated by the materials present. By transferring the X-rays in a scintillator layer into visible light, the detector records the corresponding 2D attenuation image (penetration image). The specimen is rotated stepwise recording at each position a 2D attenuation image to generate the full series of attenuation images of a $360^\circ$ rotation for a complete reconstruction of the data volume ~\cite{heinzl-2008-thesis}. Finally the series of penetration images is reconstructed to a 3D volumetric dataset of the specimen. 
While cone beam XCT can reach voxel sizes below 500 $nm$ generating high resolution volume data for comprehensive and detailed analyses, unfortunately there is still a trade-off between viewport and image resolution. The magnification reached within an XCT scan is determined by the used distances between source and specimen as well as source and detector. The magnification therefore directly influences both resolution and viewport: while higher resolutions decrease the viewport but show more details, lower resolutions allow for larger viewports and thus larger portions of the specimen.

In this work, we focus on datasets with larger viewports and lower resolutions where the individual carbon fibers (filaments) are indiscernible or barely visible. Our domain experts are mainly interested in visualizing the geometric structures in the weaving pattern of fiber bundles in endless carbon fiber reinforced composites instead of high resolution studies of individual fibers.
%of short fiber reinforced composites as presented by Weissenbock2014 et al.~\cite{Weissenbock2014}.
Figure~\ref{fig:data-char} provides an example of our targeted dataset type. It also shows the recurring fiber bundle pattern in the final CFRP laminate, called a \textit{unit cell}.
Our work is motivated from the recent progress in two interrelated fields: Firstly, CFRP components have gained wide application in industry because of its superior material and physical properties in comparison to conventional materials~\cite{Karpat2012}. Secondly, recent developments of industrial 3D X-ray computed tomography (XCT) with regard to larger detectors, larger field of views, and better resolutions opened XCT for this new application area of non destructive testing for fiber reinforced components~\cite{Schilling2005}. 
% Fiber bundles determine to a large extent the properties in terms of specific stiffness, specific strength, or durability of the final component. However, when scanning reinforced polymers using modern XCT devices, there is still a trade-off between view port and image resolution. Due to the intrinsic concept of cone beam and fan beam XCT setups, the magnification reached within a scan is determined by the used distances between source and specimen as well as source and detector. Furthermore the magnification directly influences both resolution and viewport: Higher resolutions decrease the viewport or how much of the structure can be imaged while lower resolutions allow for larger view ports.
%\begin{center}
%\includegraphics[width=0.3\textwidth]{imagesMetaTracts/crop-13-total}
%\caption{Volume rendering of a part of a dataset. Fig~\ref{fig:slice_data:a} and ~\ref{fig:slice_data:b} show image slices along the blue and red plane respectively.  }\label{fig:orig_data}
%\end{center}
%!!!!! TODO REFINE DTI PARAGRAPH !!!!!
%Diffusion Tensor Imaging tractography clustering and visualization.
%The topic of DTI visualization has seen an extensive amount of investigation by researchers.
%Advantage of two phase clustering 
%What is the  motivation for fiber-lets
%Seed point selection 
%distance between tracks 
%how well do we do in noise , may be data set 1
%!!!!! I have replaced this with the following para !!!!!
%In this work, we present a framework which interprets and advances these techniques in order to extract and visualize fiber bundles in industrial 3D X-Ray computed tomography data of woven fiber reinforced composites. The main goal of this work is to expand the state of the art in nondestructive testing of composite structures towards lower resolutions and thus larger areas of woven fabrics.
%While the importance of fiber bundles in determining the final properties of the components have  become evident,  the tools for visualization of the same have not developed at the same pace.
While fiber bundles are now understood as highly important in determining component properties, the tools for visualization of the internal structure have not developed at the same pace.
To the best of our knowledge, there is no current work that can resolve simple queries such as:
\begin{itemize}[noitemsep]
\item{ How to extract and visualize the geometric structure of a particular fiber bundle?}
\item{ How to visualize the interaction between a particular pair of fiber bundles (weaving/braiding) or a unit cell?}
\item{ Which fiber bundles show a particular orientation? }
\item{ Which fiber bundles are of the same type of yarn (i.e. which bundles show similar sizes or diameters), which is the largest or smallest fiber bundle?}
\end{itemize}

We demarcate the above queries into two parts: \textit{geometric structure} and \textit{spatial context}. Geometric structure refers to the shape, size, and orientation of a fiber bundle or a group of bundles. Spatial context refers to how two or more bundles interact with each other. It provides answers to questions such as: Are these bundles in contact at a particular position in the data? What is the relative orientations of the contacting bundles?

Providing answers to the above queries from the volume renderings of XCT datasets, or from visual inspection of particular 2D slices is non-trivial, even for experts. We present and evaluate a method which uses visualization techniques to gain insight into our data.
We interpret and advance techniques from diffusion tensor imaging to extract and visualize geometric structures from XCT data of the woven carbon fiber reinforced composites. The main goal of this work is to expand the state of the art in non-destructive testing through visualization of composite structures in \textit{lower resolutions} and thus \textit{larger areas} of woven fabrics.
\begin{figure}[tb]
	\centering
	%\includegraphics[width=0.45\textwidth]{imagesMT2014/algo.png}
	%\includegraphics[width=0.5\textwidth,  trim = 0mm 300mm 0mm 0mm, clip,]{imagesMT2014/image_algo}
	\includegraphics[width=\linewidth]{images_pvis/workflow.pdf}
	\caption{Flow chart of the MetaTracts approach for fiber bundle extraction}
	\label{fig:flowchart}
\end{figure}
Figure~\ref{fig:flowchart} shows an over view of our approach. Starting from the XCT data, we first generate MetaTracts. Then we cluster the MetaTracts to generate and visualize fiber bundles. For a detailed description of the MetaTracts generation process see Sec.~\ref{sec:approach}. For the description of plausible fiber bundle generation see Sec.~\ref{subsec:fiber-bundles}. For a discussion of the visualizations we create for our domain experts see Sec.~\ref{sec:vis}. Sec.~\ref{sec:results} describes our experimental results, Sec.~\ref{sec:param_choices} presents our parameter choices and  Sec.~\ref{sec:user_eval} summarizes our user evaluations.

@@

Diffusion Magnetic Resonance Imagining (dMRI, also referred to as Diffusion Tensor Imaging (DTI)) has gained popularity in medical diagnosis within recent years. Its main clinical application is found in the study and treatment of neurological disorders. DTI may reveal abnormalities in white matter fiber structure and is used for visualizing the organization of white fibers in the human brain and  brain connectivity.
A variety of algorithms have been proposed aiming for generating fiber-tract trajectories. In general, these reconstructions of fiber trajectories are  clustered in bundles which are expected to be related anatomically or spatially. 
We broadly divide the related work on DTI into two parts that are of immediate relevance to our proposed solution: fiber tracking and fiber clustering. We discuss  methods for analyzing the second order local structure and review the current state of the art in the visual analysis of fiber reinforced composites.

\subsection {Fiber Tracking in Diffusion Tensor Imaging}
\label{subsec:fiberEx} 
A basic assumption in DTI analysis is that the principal eigenvector of the diffusion tensor is parallel to the underlying dominant fiber direction in each image voxel~\cite{Basser2002, Mori1999, Mori2002}. Continuous tracts are created by propagating virtual particles along fiber directions until they reach some termination criterion. 
This is usually done by solving a Runge-Kutta integration (typically second or fourth order). The principal diffusion direction at each discrete location is interpolated to form a continuous velocity field. 
Because decisions are made locally, these methods perform poorly in noisy regions and often generate small fibers. Basser et al.~\cite{Basser2002} proposed that white matter tracts could be represented as 3D curves in space. They showed that numerical methods could be used to follow fibers and fiber bundles and to generate tracts in human brain data. 
Mori et al.~\cite{Mori1999,Mori2002} divided reconstruction techniques into line-propagation or energy minimization techniques. In line propagation approaches, trajectories are computed based on local neighborhoods and in energy minimization approaches the most  favorable trajectory connecting two given endpoints is selected.

\subsection {Fiber Clustering in Diffusion Tensor Imaging}
\label {subsec:fiberClus}
In DTI, similarity measures are used to group fiber tracts into bundles. Extensive research has been done on automatic DTI fiber clustering methods \cite{Brun2004,Brun2003,Corouge2004,westinMEDIA02,Zhang2008}. 
Clustering requires a suitable proximity measure and a method for grouping ``close" fibers.
Pairwise proximity measures include endpoint distances~\cite{Brun2003} and mean of the closest distances between points on two fibers~\cite{Corouge2004}. Zhang et al.~\cite{Zhang2008} introduced a thresholded version of the of mean of closest distances, so that fibers that are close for certain distance and then diverge, are clustered separately. Brun et al.~\cite{Brun2004} uses normalized cuts along with a pairwise distances  measure computed using a 9-D fiber shape descriptors. The choice of the clustering algorithm can be broadly divided into those approaches using hierarchical clustering~\cite{Moberts2005, Zhang2008} and those using spectral clustering~\cite{ Brun2004,jonasson2005, ODonnell2007}.

Brun et al.~\cite{Brun2003} described how a spectral non-linear dimensionality reduction technique,  Laplacian eigenmaps (Belkin and Niyogi~\cite{Belkin01}), can be applied to the problem of organizing fiber tracts data. The key notion of the Laplacian eigenmaps algorithm is to represent the underlying data as a graph. Each node represents a data point and the edges connect neighboring data points. An eigenvalue problem is solved to represent the data in a lower dimensional space while preserving the local graph structure. In the case of fiber bundles, the individual points are fiber tracts. In the ideal case fiber tracts that belong to the same bundle must remain ``close" to each other in the lower dimensional space. Westin et al.~\cite{westinMEDIA02} also use spectral clustering on a Hausdorff distance measure defined as the maximum of point wise minimum distances between two fibers. Jonasson et al.~\cite{jonasson2005} run k-means clustering on the eigenvectors of the affinity matrix defined as the co-occurrence of fibers in the data.
%The clustering methods have also been evaluated to be generally accurate in capturing fiber bundles as seen in Moberts et al.~\cite{Moberts2005} and Corouge et al.~\cite{Corouge2004}.

The agglomerative hierarchical clustering method \cite{DudaHartStork01} has gained popularity for proximity based fiber segregation (Zhang et al.~\cite{Zhang2008}, Corouge et al.~\cite{Corouge2004}). 
These approaches build on the assumption that proximity measures that compare DTI fiber trajectories can also represent anatomical relationships. An agglomerative hierarchical clustering method starts with each data point/fiber in an individual cluster. At each stage of the algorithm the two most similar clusters are merged based on some criterion. The two basic cluster similarity measures are single-link and complete-link. With the single link measure the distance between the clusters is the distance between the closest pair of items. 
Moberts et al.~\cite{Moberts2005} implemented several distance measures in their evaluation of fiber clustering and concluded that clustering methods are generally accurate in capturing fiber bundles. 
%Mean of closest distances performs better than closest point distance computations or Hausdorff distance measures and hierarchical clustering using single-link measure gave the best results. 
There are a number of difficulties in hierarchical clustering. First, computing all pairs distances for tracts to generate the distance matrix is time consuming. Second, a ``correct'' distance measure to compare tracts must be chosen. Third, hierarchical clustering is best suited for similar length fibers.
Spectral methods are also hindered by long matrix computations.

\subsection{Second order local structure}
%!!!!! TODO REFINE SECTION !!!!!
Unlike DTI, we do not have diffusion tensor data. Instead, we have a scalar volume with tubular structures embedded in them. Analyzing curvilinear structures in volumetric images has been utilized for a variety of purposes including center line extraction~\cite{Bouix2005} and vascular image enhancement~\cite{Frangi1998,Sato1997}. Frangi et al.~\cite{Frangi1998} introduced a method based on studying the eigenvalues of the  Hessian matrix specifically for the purposes of developing vessel enhancement filters.

\subsection {Visual Analysis and Modeling of Fiber Reinforced Composites}
The approaches presented in visualization and analysis of composites mainly focus on individual objects such as fiber extraction from high resolution data where the individual fibers are clearly discernible. Fritz et al.~\cite{Fritz2009} proposed interactive workflows%for non destructive testing practitioners 
to explore and quantify steel fibers in reinforced sprayed concrete. %This approach allows analyzing fiber orientations based on direction transfer functions. 
Salaberger et al.~\cite{Salaberger2011} introduced a pipeline to extract and characterize individual fibers of fiber reinforced composites.%They encode the extracted fibers as color-coded line segments in 3D and visually identifying fibers with similar orientations.
Recently, Weissenbock et al.~\cite{Weissenbock2014} introduced a system for interactive exploration and analysis of fibers in fiber reinforced polymers. 
Lomov et al.~\cite{Lomov2010} discusses the problems and current available solutions in geometric modeling of three dimensional composites.
%Modeling of the composites, first starts with establishing the topology of the structure, which translates to answering if a particular bundle is in contact with another at a particular position? The second step builds the geometry of the model, answering questions relating to placement of bundles in space their orientations and dimensions.

%Tools based on image analysis, fiber tracking and clustering into bundles in DTI and their subsequent visualization through stream tubes and coloring schemes have received extensive research. Recently, similar techniques have been applied to fiber tracing in composite materials. 
%To the best of our knowledge there is no approach focusing on direct extraction of fiber bundle structures from low resolution XCT of composite fibers.

@@

\section {Extracting MetaTracts }
\label{sec:approach}
Extracting the MetaTracts consists of two  major steps. First we extract a local orientation vector at each grid location (Sec.~\ref{subsec:rh}). Second, we compute coarse poly-cylinders called MetaTracts which traverse the data (Sec.~\ref{subsec:mtprop}).

@@

The input to the first stage of our pipeline, is the original scalar dataset in a uniform lattice grid in $\mathbb{R}^3$. As output, each grid location becomes associated with an unit vector representing the local orientation and a real value [0,1] that represents a measure of the reliability of the local orientation. We approximate the local orientation by eigenvalue analysis of the Hessian matrix computed at each voxel.
The Hessian matrix represents the local curvature. The eigen decomposition of the Hessian matrix gives the eigenvectors which describe the local second-order structure of the image. The eigenvector corresponding to the smallest eigenvalue (also referred as the principal direction) gives the direction along which the curvature is smallest. This direction coincides with the direction of the tubular structure.
Frangi et al.~\cite{Frangi1998} introduced a process that searches for geometric structures which are tubular. They define a measure based on two geometric ratios of the second order ellipsoid given by the local Hessian matrix to measure the ``vesselness'' criterion.  In order to determine reliable Hessians, we compute 
 the same metric. We include their work here for completeness and direct the reader to ~\cite{Frangi1998} for details. Let $\lambda_{K}$ be the eigenvalue with the $K^{th}$ smallest magnitude. Here, $|{\lambda}_{1}| \leq| {\lambda}_{2}|\leq| {\lambda}_{3}| $ are the eigenvalues of the Hessian matrix. Specifically, a pixel belonging to a vessel region will have small $\lambda_{1}$ ($|\lambda_{1}|\approx 0$) and $\lambda_{2}$, $\lambda_{3}$ of large magnitude and of equal sign($|\lambda_{1}| \ll |\lambda_{2}|$ and $|\lambda_{2}|\approx |\lambda_{3}|$). The sign indicates if the vessel is bright in a dark background or dark in a bright background. In our case the individual fibers are bright ($\lambda_2,\lambda_3 < 0$). The following measures are defined in~\cite{Frangi1998}.  
\begin{equation}\label{RA}
\mathcal{R_{A}}=\frac{\textrm{Largest  Cross Section}\big/ \pi}{{\textrm{Largest Axis Semi-length}}^{2}}=\frac{|\lambda_{2}|}{|\lambda_{3}|}
